### 2020-03-02 doc/icfp20/counterpoint.tex
proof read sections 1 - 3.2

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which is required to start with a tonic (the first note of a scale;
e.g., C in the case of C major) and consist only of whole notes.
Figure~\ref{fig:fs} shows an example of first species counterpoint.
The lower line is the cantus firmus, which we excerpt from a German
song called \emph{Froschgesang} (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

\begin{alltt}
PitchInterval : Set
PitchInterval = Pitch \(\times\) Interval

first : PitchInterval
first = (c 5 , per8)

@@

The beginning of the music should express perfection.
As we saw in Section~\ref{sec:music}, there are three intervals that
are classified as perfect: the 1st (usually called the \emph{unison}),
t5th, and 8th.
The first interval of the music must then be one of these intervals.
In our formalization, we implement this rule as the
\texttt{checkBeginning} function, which reports an error
\texttt{not158 i} when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we turn the error into an option
value by wrapping it around the \texttt{just} constructor.

@@

The middle bars of the music should maintain harmonical consonance
and independence of melodic lines.
As we saw before, consonant intervals include the unison,
3rd\footnote{We use ``3rd'' to mean both the major and minor variants
of the 3rd interval, and similarly for the 6th.}, 5th, 6th, and 8th,
and among these, the unison is clearly an obstacle to distinguishing
between the two lines of music.
Therefore, the middle bars must consist of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors correponding to the occurrences of dissonant
intervals and unisons.

@@

intervalCheck : PitchInterval \(\rightarrow\) Maybe IntervalError
intervalCheck (p , i) with isConsonant i | isUnison i
... | false | _    = just (dissonant i)
... | _     | true = just (unison p)
... | _     | _    = nothing

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As Figure~\ref{fig:motioncadence} left shows, there are four kinds of motion:
\emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction),
and \emph{oblique} (one line plays the same note).
It is known that approaching a perfect interval by parallel or similar
motion tends to fuse the two lines of melody into one.
To rule out such motion, we define the \texttt{checkMotion}
function, which inspects all pairs of two adjacent intervals (using an
auxiliary function \texttt{pairs}) and returns a list of erroneous patterns.

@@

Using the encoding of music and rules we have seen so far, we define
\texttt{FirstSpecies}, a record type inhabited by correct counterpoint.
The first three fields of this record type represent the first, middle,
and last bars, respectively.
The last four fields stand for the proofs that the music satisfies all
the required properties, that is:

\begin{itemize}
  \item The beginning is a perfect consonance
  \item The middle bars have no dissonant interval or unison
  \item The middle bars have no perfect interval approached by
    parallel or similar motion
  \item The ending is a unison or an octave approached by a cadence
\end{itemize}

\noindent Recall that the functions representing these rules return
either an option value or a list of found errors.
Therefore, the correctness proofs either take the form
\texttt{func args $\equiv$ nothing}, or look like
\texttt{func args $\equiv$ []}.

@@

With this record type, we can show that the counterpoint in
Figure~\ref{fig:fs} is correct, since the music is an inhabitant of
the \texttt{FirstSpecies} type.

@@

\begin{alltt}
PitchInterval2 : Set
PitchInterval2 = Pitch \(\times\) Interval \(\times\) Interval

first2 : PitchInterval
first2 = (c 5 , per5)

middle2 : List PitchInterval2
middle2 =
  (d 5 , min3 , per5) :: (e 5 , min3 , min6) :: (f 5 , maj3 , aug4) ::
  (e 5 , min6 , min3) :: (d 5 , min3 , maj6) :: []

@@

Weak beats are constrained in a looser manner than strong beats.
First, they are allowed to be dissonant if they are created by a
\emph{passing tone}, i.e., a note in the middle of two step-wise
motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
Second, they are allowed to be the unison if they are left by step in
the opposite direction from their approach (as in bars 5-6 of
Figure~\ref{fig:ss}).
% We implement these rules as the \texttt{checkWeakBeats} function,
% which checks the validity of every three successive beats in the
% middle bars.

@@

The last interval must again be the unison or 8th, preceded by an
appropriate interval that constitutes a cadence structure.

@@

Now we define \texttt{SecondSpecies}, a record type inhabited by
correct second species counterpoint.
As in \texttt{FirstSpecies}, we have three fields holding the musical
content, followed by five fields carrying the proofs of the required
properties disucssed above\footnote{
  As the reader would have imagined, each \texttt{checkXXX} function
  corresponds to one rule that must be satisfied by the counterpoint.
  The auxiliary function \texttt{secondPitch} in \texttt{weakBeatsOk}
  extracts the counterpoint note of the last interval, and
  \texttt{expandPitchIntervals2} turns a list of \texttt{PitchInterval2}s
  into a list of \texttt{PitchInterval}s.
}.

>>>>

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which is required to start with a tonic (the first note of a scale,
such as C in C major) and consist only of whole notes.
Figure~\ref{fig:fs} shows an example of first species counterpoint.
The lower line is the cantus firmus, which we excerpt from a German
song called \emph{Froschgesang} (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

\begin{alltt}  
first : PitchInterval -- Pitch \(\times\) Interval
first = (c 5 , per8)

@@

The beginning of the music should express perfection.
As we saw in Section~\ref{sec:music}, there are three intervals that
are classified as perfect: the 1st (usually called the \emph{unison}),
5th, and 8th.
The first interval of the music must then be one of these intervals.
In our formalization, we implement this rule as the
\texttt{checkBeginning} function, which reports an error
\texttt{not158 i} when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we turn the error into a
\texttt{Maybe} value by wrapping it around the \texttt{just} constructor.

@@

The middle bars of the music should maintain harmonical consonance
and independence of melodic lines.
As we saw before, consonant intervals include the unison,
3rd\footnote{We use ``3rd'' to mean both the major and minor variants
of the 3rd interval, and similarly for the 6th.}, 5th, 6th, and 8th,
and among these, the unison is clearly an obstacle to distinguishing
between the two lines of music.
Therefore, the middle bars must consist only of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors correponding to the occurrences of dissonant
intervals and unisons.

@@

intervalCheck : PitchInterval \(\rightarrow\) Maybe IntervalError
intervalCheck (p , i) with isConsonant i | isUnison i
intervalCheck (p , i) | false | _    = just (dissonant i)
intervalCheck (p , i) | _     | true = just (unison p)
intervalCheck (p , i) | _     | _    = nothing

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As Figure~\ref{fig:motioncadence} left shows, there are four kinds of motion:
\emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction by
different distances), and \emph{oblique} (one line plays the same note).
It is known that approaching a perfect interval by parallel or similar
motion tends to fuse the melodic lines into one.
To rule out such motion, we define the \texttt{checkMotion}
function, which inspects all pairs of two adjacent intervals (using an
auxiliary function \texttt{pairs}) and returns a list of erroneous patterns.

@@

Using the encoding of music and rules we have seen so far, we define
\texttt{FirstSpecies}, a record type inhabited by correct first species
counterpoint.
The first three fields of this record type represent the first, middle,
and last bars, respectively.
The last four fields stand for the proofs that the music satisfies all
the required properties.

@@

With this record type, we can show that the counterpoint in
Figure~\ref{fig:fs} is correct, since the music is an inhabitant of
\texttt{FirstSpecies}.

@@

\begin{alltt}
first2 : PitchInterval
first2 = (c 5 , per5)

middle2 : List PitchInterval2 -- Pitch \(\times\) Interval \(\times\) Interval
middle2 =
  (d 5 , min3 , per5) :: (e 5 , min3 , min6) :: (f 5 , maj3 , aug4) ::
  (e 5 , min6 , min3) :: (d 5 , min3 , maj6) :: []

@@

Weak beats are allowed to be dissonant if they are created by a
\emph{passing tone}, i.e., a note in the middle of two step-wise
motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
They may also be the unison if they are left by step in the opposite
direction from their approach (as in bars 5-6 of Figure~\ref{fig:ss}).
% We implement these rules as the \texttt{checkWeakBeats} function,
% which checks the validity of every three successive beats in the
% middle bars.

@@

The last interval must be the unison or 8th, preceded by an
appropriate interval that constitutes a cadence structure.

@@

Now we define \texttt{SecondSpecies}, a record type inhabited by
correct second species counterpoint.
As in \texttt{FirstSpecies}, we have three fields holding the musical
content, followed by five fields carrying the proofs of the required
properties disucssed above\footnote{
  The auxiliary function \texttt{secondPitch} in \texttt{weakBeatsOk}
  extracts the counterpoint note of a given interval, and
  \texttt{expandPitchIntervals2} turns a list of \texttt{PitchInterval2}s
  into a list of \texttt{PitchInterval}s.
}.

### 2020-03-02 doc/icfp20/intro.tex
proof read sections 1 - 3.2

Full dependent types allow expression of predicate logic, and it is
tempting to take a standard textbook on music theory such as
\citet{piston-harmony} or \citet{aldwell2018harmony} and formalize it
in type theory.  As a first step we start with the modest task of
expressing a small, relatively strict rule set known as species
counterpoint~\cite{fux-cp}, intended for combining interdependent
melody lines to produce a pleasant-sounding result.  Roughly speaking,
what we do is to write a custom musical type checker in Agda.
Compared to the previous work by \citet{cong-cp}, where rules are
expressed directly as Agda types, our approach makes it easier to
describe fine-grained rules, produce readable error messages, and add
or drop rules depending on the circumstances.

@@

The rest of this paper structured as follows.  Section~\ref{sec:music}
provides a brief glossary of musical terms, to prepare the
reader with no familiarity in music theory.  Sections~\ref{sec:cp} and
\ref{sec:harmony} describe our formalization of counterpoint and
harmony, focusing on the composable and modular aspects of the
proposed approach.  Lastly, Section~\ref{sec:conclusion} concludes
the paper with future perspectives.

>>>>

Full dependent types allow expression of predicate logic, and it is
tempting to take a standard textbook on music theory such as
\citet{piston-harmony} or \citet{aldwell2018harmony} and formalize it
in type theory.  As a first step towards this goal, we start with the modest task of
expressing a small, relatively strict rule set known as species
counterpoint~\cite{fux-cp}, intended for combining interdependent
melody lines to produce a pleasant-sounding result.  Roughly speaking,
what we do is to write a custom musical type checker in Agda.
Compared to the previous work by \citet{cong-cp}, where rules are
expressed directly as Agda types, our approach makes it easier to
describe fine-grained rules, produce readable error messages, and add
or drop rules depending on the circumstances.

@@

The rest of this paper structured as follows.  Section~\ref{sec:music}
introduces basic musical concepts as well as their Agda representation.
Sections~\ref{sec:cp} and \ref{sec:harmony} describe our formalization
of counterpoint and harmony, focusing on the composable and
modular aspects of the proposed approach.  Lastly,
Section~\ref{sec:conclusion} concludes the paper with future perspectives.

### 2020-03-02 doc/icfp20/music.tex
minor fixes
proof read sections 1 - 3.2

\section{Musical Preliminaries}
\label{sec:music}

\paragraph{Pitches}
A \emph{pitch} is a value denoting how high or low a sound is.
In our implementation, pitches have type \texttt{Pitch} and take the
form \texttt{name octave}, where \texttt{name} is a name ranging
over \texttt{c}, \texttt{d}, \texttt{e}, etc. and \texttt{octave} is a
natural number denoting which octave the pitch belongs to.
For instance, the middle C is represented as \texttt{c 5}.

@@

\paragraph{Duration}
\emph{Duration} denotes an unspecified unit of time during which a
sound or silence lasts.
In our implementation, duration (of type \texttt{Duration} is simply
a natural number, denoting the relative length within a bar (such as
whole and half).
When the music is played at a specific tempo, the number is 
multiplied by a velocity and turned into an absolute length.

@@

\paragraph{Notes}
Combining pitches and duration gives us \emph{notes}.
In our implementation, we represent notes as a datatype \texttt{Note}
with two constructors: \texttt{tone} for notes with sound and
\texttt{rest} for those without.

@@

\emph{Intervals} are a concept denoting the difference in pitch
between two notes.
There are 13 kinds of interval within an octave, and these intervals
can be classified from several different perspectives: (i) major or
minor; (ii) consonant or dissonant; and (iii) perfect or imperfect.
We define intervals as a datatype \texttt{Interval}, where each
constructor represents one of the 13 intervals.

\begin{figure}[h]
  \includegraphics[width=12cm]{fig/interval.png} \\
  \begin{flushleft}
    \begin{footnotesize}
      \hspace{1.45cm} \texttt{per1 \hspace{0.5mm} min2  \hspace{2.5mm}
        maj2 \hspace{1.2mm} min3 \hspace{0.5mm} maj3 \ per4
        \hspace{1mm}aug4
        \hspace{0.5mm}  per5\hspace{1.2mm}  min6\hspace{1.6mm} maj6
        \hspace{1.6mm}min7\hspace{1.2mm} maj7\hspace{1.2mm} per8} \\
      consonant? \hspace{1.5mm} yes \hspace{4.2mm} no \hspace{6.5mm}  no
      \hspace{4.7mm} yes \hspace{3.8mm}  yes \hspace{3.7mm} no \hspace{3.5mm}
      no \hspace{4.3mm} yes \hspace{3.8mm} yes \hspace{3.2mm} yes
      \hspace{3.7mm} no \hspace{4mm} no \hspace{3.8mm} yes
    \end{footnotesize}
  \end{flushleft}
\end{figure}

>>>>

\newcommand{\Interval}{
  \begin{figure}
  \includegraphics[width=12cm]{fig/interval.png} \\
  \begin{flushleft}
    \begin{footnotesize}
      \hspace{1.45cm} \texttt{per1 \hspace{0.5mm} min2  \hspace{2.5mm}
        maj2 \hspace{1.2mm} min3 \hspace{0.5mm} maj3 \ per4
        \hspace{1mm}aug4
        \hspace{0.5mm}  per5\hspace{1.2mm}  min6\hspace{1.6mm} maj6
        \hspace{1.6mm}min7\hspace{1.2mm} maj7\hspace{1.2mm} per8} \\
      consonant? \hspace{1.5mm} yes \hspace{4.2mm} no \hspace{6.5mm}  no
      \hspace{4.7mm} yes \hspace{3.8mm}  yes \hspace{3.7mm} no \hspace{3.5mm}
      no \hspace{4.3mm} yes \hspace{3.8mm} yes \hspace{3.2mm} yes
      \hspace{3.7mm} no \hspace{4mm} no \hspace{3.8mm} yes
    \end{footnotesize}
  \end{flushleft}
  \caption{13 Kinds of Intervals}
  \label{fig:interval}
 \end{figure}
}

\section{Basic Musical Terms and Their Representation in Agda}
\label{sec:music}

\paragraph{Pitches}

A \emph{pitch} tells us how high or low a sound is.
We represent pitches as natural numbers, but throughout the paper,
we use a more readable notation \texttt{name octave}, where
\texttt{name} ranges over \texttt{c}, \texttt{d}, \texttt{e}, etc.,
and \texttt{octave} denotes which octave the pitch belongs to.
For instance, the middle C is represented as \texttt{c 5}.

@@

\paragraph{Duration}

\emph{Duration} denotes an unspecified unit of time during which
a sound or silence lasts.
We represent duration as a natural number, and use this value to
calculate the absolute length when the music is played at a specific
tempo.
As an example, the duration constant \texttt{whole} has value
\texttt{16} and corresponds to 4 seconds if the tempo is 60 beats
per minute.

@@

\paragraph{Notes}

Combining pitches and duration gives us \emph{notes}.
We represent notes as a datatype \texttt{Note} with two constructors:
\texttt{tone} for notes with sound and \texttt{rest} for those without.
For example, \texttt{tone whole (c 5)} is a whole note whose sound
is the middle C.

@@

\Interval

An \emph{interval} represents the difference in pitch between
two notes.
There are 13 kinds of interval within an octave
(Figure~\ref{fig:interval}, and these intervals can be classified from
several different perspectives:
(i) major or minor; (ii) consonant or dissonant; and (iii) perfect
or imperfect.
We define intervals as a datatype \texttt{Interval}, where each
constructor represents one of the 13 intervals.
